---
title: "(Sliced) Wasserstein - Gaussian process regression workflow"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
    fig_align: "center"
    dev: svg
vignette: >
  %\VignetteIndexEntry{(Sliced) Wasserstein - Gaussian process regression workflow}           
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Goal
This vignette illustrates the main use of the `slicer` package
by means of a toy example.
Empirical distributions are generated from bivariate Gaussian mixtures 
with four isotropic components laid out symmetrically around the origin.
These distributions differ from one another in that they are rotated
counterclockwise by an angle $\omega \in [0, \pi/2)$.
The goal of the analysis is to retrieve the specific angles by which
the distributions have been rotated.

## Functions
Let's load `slicer`.
```{r setup}
library(slicer)
```

To generate the input distributions, we define `generate_data()`:
```{r}
generate_data <- function(n = 200, a = 5, angle = 0, Sigma = diag(1, 2)) {
  mus <- cbind(c(a, 0), c(0, a), c(-a, 0), c(0, -a))
  R <- rbind(
    c(cos(angle), -sin(angle)),
    c(sin(angle), cos(angle))
  )
  cluster <- sample(1:4, size = n, replace = TRUE)
  d <- matrix(0, nrow = n, ncol = 2)
  for (i in 1:4) {
    d[which(cluster == i), ] <- MASS::mvrnorm(
      sum(cluster == i), mu = mus[, i], Sigma = Sigma
    )
  }
  d %*% t(R)
}
```

The plots below show two empirical distributions,
once using a rotation angle of $\omega = \pi/16$ and once using 
one of $\omega = 3\pi/8$.

```{r}
#| fig.width = 9
set.seed(2026) # for reproducibility

op <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
generate_data(angle = pi/16) |> 
  plot(xlab = "first dimension", ylab = "second dimension", asp = 1)
generate_data(angle = 3*pi/8) |> 
  plot(xlab = "first dimension", ylab = "second dimension", asp = 1)
par(op)
```

## Data generation
We generate 60 matrices (40 for training, 20 for testing) 
containing empirical distributions
with the rotation angles sampled uniformly from $[0, \pi/4)$.
Each empirical distribution is based on 200 data points.

```{r}
sample_size <- 200
N_train <- 40
N_test  <- 20
angle_range <- c(0, pi/4)

angles <- runif(N_train + N_test, angle_range[1], angle_range[2])

distributions <- vector("list", N_train + N_test)
for (i in seq_len(N_train + N_test)) {
  distributions[[i]] <- generate_data(angle = angles[i])
}
```

## Distances
We can use `compute_all_distances()` to compute pairwise distances
between the input distributions.
If we supply the optional vector `test_idx` with the test set indices,
the pairwise distance computations between test objects 
(which are not needed for obtaining predictions) are skipped.
To compute **sliced Wasserstein distances**, we need to first generate
$\boldsymbol{\theta_1}, \dots, \boldsymbol{\theta_L}$ 
sampled uniformly at random from the unit sphere;
`generate_directions()` takes care of this.
We'll use $L = 25$ and set $d = 2$ since we're working in two dimensions.
If we're only interested in the estimated sliced Wasserstein distances,
we can set `keep_projections` to `FALSE`.

```{r}
thetas <- generate_directions(L = 25, d = 2)
sw_distances <- compute_all_distances(distributions, thetas, verbose = FALSE,
    keep_projections = FALSE, test_idx = N_train + seq_len(N_test))
```

The $60 \times 60$ matrix `sw_distances` contains the _squared_ (estimated)
sliced Wasserstein distances among the forty training objects and between
the training objects and the twenty test objects.
The pairwise distances among the test objects are set to `NA`:

```{r, eval = FALSE}
sw_distances # output not shown
```


We can also compute (normal) **Wasserstein distances** between 
the distributions when they are projected along certain dimensions.
For instances, to compute the pairwise Wasserstein distances along
the first margin and along the second margin, we set the 
projection directions to 
$\boldsymbol{e_1} = (1, 0)^{\top}$ and 
$\boldsymbol{e_2} = (0, 1)^{\top}$.
We also set `keep_projections = TRUE`, which will cause the function to
output a list of two matrices with _squared_ Wasserstein distances:
one for each margin.

```{r}
marginal_distances <- compute_all_distances(distributions, diag(1, 2), 
    verbose = FALSE, keep_projections = TRUE, 
    test_idx = N_train + seq_len(N_test))
str(marginal_distances)
```

## Gaussian process models with tuned hyperparameters
The function `fit_gpr()` takes a single matrix with squared pairwise 
distances and uses it as input to a Gaussian process regression model
with (by default) a Gaussian RBF kernel:
$$k_{\textrm{RBF}}(x, x') = s^2\exp\left(-\frac{d^2(x, x')}{2t^2}\right)$$,
where $s^2 > 0$ is a scaling factor, $t > 0$ is the length-scale,
and $d(x, x')$ is the distance between the inputs $x$ and $x'$.

The model's and the kernel's hyperparameters
are tuned using the training data by minimising the negative
marginal log-likelihood. 

```{r}
sw_fit <- fit_gpr(sw_distances,
  training_idx = seq_len(N_train),
  test_idx = N_train + seq_len(N_test),
  y_train = angles[seq_len(N_train)], 
  y_test = angles[N_train + seq_len(N_test)],
  verbose = TRUE)
```

The output includes predictions for the test data,
the covariannce for the predictions
(if the true test outcomes were provided), 
and the estimated hyperparameters.

```{r}
str(sw_fit)
plot(angles[N_train + seq_len(N_test)], sw_fit$test_predictions,
     xlab = "true test outcomes", ylab = "predicted test outcomes", asp = 1)
```

Since we didn't compute the distances between the test inputs, 
only the predictions variances are reported (the main diagonal of `test_variance`):
```{r}
diag(sw_fit$test_variance)
```
Had we also computed the distances between the test inputs,
`test_variance` would have also included the covariances.

The function `fit_gpr()` can also be used when multiple matrices 
with squared distances are provided.
Now, estimated length-scale and (kernel) variance hyperparameters
are provided for the Gaussian RBF corresponding to each squared distance matrix.

```{r}
marginal_fit <- fit_gpr(marginal_distances,
  training_idx = seq_len(N_train),
  test_idx = N_train + seq_len(N_test),
  y_train = angles[seq_len(N_train)], 
  y_test = angles[N_train + seq_len(N_test)],
  verbose = TRUE)
str(marginal_fit)
plot(angles[N_train + seq_len(N_test)], marginal_fit$test_predictions,
     xlab = "true test outcomes", ylab = "predicted test outcomes", asp = 1)
```

We can combine the marginal and sliced Wasserstein distances into a list with
three distance matrices, too:

```{r}
total_fit <- fit_gpr(list(sw_distances, marginal_distances[[1]], marginal_distances[[2]]),
  training_idx = seq_len(N_train),
  test_idx = N_train + seq_len(N_test),
  y_train = angles[seq_len(N_train)], 
  y_test = angles[N_train + seq_len(N_test)],
  verbose = TRUE)
str(total_fit)
plot(angles[N_train + seq_len(N_test)], total_fit$test_predictions,
     xlab = "true test outcomes", ylab = "predicted test outcomes", asp = 1)
```

Parallel processing can be enabled using the `cores` parameter:

```{r}
total_fit <- fit_gpr(list(sw_distances, marginal_distances[[1]], marginal_distances[[2]]),
  training_idx = seq_len(N_train),
  test_idx = N_train + seq_len(N_test),
  y_train = angles[seq_len(N_train)], 
  y_test = angles[N_train + seq_len(N_test)],
  runs = 50L, cores = 2L)
str(total_fit)
```

Instead of the Gaussian RBF kernel, the Matérn kernels with
smoothness parameter $\nu = 0.5, 1.5, 2.5$ can be used.
We can use a different kernel for each distance matrix;
the total kernel used is then the conical combination of these kernels.
For instance, the following fit uses the Matérn kernel with $\nu = 2.5$
for the sliced Wasserstein distances,
the Matérn kernel with $\nu = 0.5$ for the Wasserstein distances along the
first axis, and the Matérn kernel with $\nu = 1.5$ for the Wasserstein
distances alogn the second axis.

```{r}
total_fit_matern <- fit_gpr(list(sw_distances, marginal_distances[[1]], marginal_distances[[2]]),
  training_idx = seq_len(N_train),
  test_idx = N_train + seq_len(N_test),
  y_train = angles[seq_len(N_train)], 
  y_test = angles[N_train + seq_len(N_test)],
  kernels = c("matern25", "matern05", "matern15"),
  runs = 50L, cores = 2L)
str(total_fit_matern)
```

Finally, a GPR's **negative log predictive density** (NLPD) for some test data
can be obtained like so:
```{r}
nlpd_gpr(total_fit_matern, angles[N_train + seq_len(N_test)])
```

The warning message occurs because we didn't compute the distances among the
test inputs themselves. As a result, the kernel matrices don't specify
the covariance among the test inputs, and the NLPD assumes that they are zero.
If you remove the `test_idx` parameter setting when running `compute_all_distances()`
and refit the models, this warning will disappear.
